---
title: ""
editor: 
  mode: source
format: 
FORMAT_INPUT
---

```{r}
#| include: false
library(ggplot2)
library(downloadthis)
library(plyr)
library(gt)
library(dataRetrieval)
library(R8WD)

Tribal_org    <- c('REPLACE_THIS_TEXT')
start_date    <- c('REPLACE_START_DATE')
end_date      <- c('REPLACE_END_DATE')
params_to_use <- c('REPLACE_PARAMS')
```

```{r}
#| echo: false
#| output: false
#| include: false

tribeData2 <- getWQP(
       organization       = Tribal_org,
       characteristicName = params$params[params_to_use],
       startDate          = start_date,
       endDate            = end_date,
       multiplier         = 0.5)

### homogenize parameter names
tribeData2$CharacteristicName <- params$new_param[match(tribeData2$CharacteristicName, params$params)]
### homogenize organization names.
### all information is interpreted as coming from a single organization
if (length(unique(tribeData2$OrganizationIdentifier) > 1)) {
  ### identify most abundant names and convert all to those
  tribeData2$OrganizationFormalName <- names(which.max(table(tribeData2$OrganizationFormalName)))
  tribeData2$OrganizationIdentifier <- names(which.max(table(tribeData2$OrganizationIdentifier)))
}

allSummary <- summarizeQC(data = tribeData2)

### identify parameters collected on sondes (no expectation of blank QC data)
sonde_parameters <- c('DO', 'pH', 'Temperature', 'Turbidity', 'Specific conductivity', 'Conductivity')
included_params <- paste0(unique(tribeData2$CharacteristicName), collapse = ', ')
included_params <- sub(",([^,]*)$",", and\\1", x = included_params)

params_not_found <- unique(params$new_param[params_to_use])[!(unique(params$new_param[params_to_use]) %in% unique(tribeData2$CharacteristicName))]
if (length(params_not_found) == 0) {
  params_not_found_message <- 'Available parameters cover the entire set of core parameters sought for this report.' 
} else if (length(params_not_found) == 1) {
  params_not_found_message <- paste0('The only core parameter sought for this report that was not available in WQP was ', params_not_found, '. EPA recommends that data for all core parameters are collected and uploaded to WQP.')
} else {
  params_not_found <- paste0(params_not_found, collapse = ', ')
  params_not_found <- sub(",([^,]*)$",", and\\1", x = params_not_found)
  params_not_found_message <- paste0('The core parameters sought for this report that were not available in WQP were ', params_not_found, '. EPA recommends that data for all core parameters are collected and uploaded to WQP.')
}
```


```{r}
#| include: false
tribal_logo_filename <- list.files(path = file.path(system.file('extdata', package = 'R8WD'), 'seals'), pattern = paste0('^', Tribal_org))
if (length(tribal_logo_filename) == 0) {
  img_path <- file.path(system.file('extdata', package = 'R8WD'), 'seals', 'epa_seal.png') # have two EPA logos if there's no organization-specific logo
} else {
  img_path <- file.path(system.file('extdata', package = 'R8WD'), 'seals', tribal_logo_filename[1])
}

# img_path_epa <- system.file('extdata/seals', 'epa_seal.png', package = 'R8WD')
img_path_epa <- file.path(system.file('extdata', package = 'R8WD'), 'seals', 'epa_seal.png')
```


## Quality Control Data Assessment Report for the `r allSummary$data$OrganizationFormalName[1]` CWA106 Program

::: {layout="[[25,-50, 25],]"}
![](`r img_path_epa`)

![](`r img_path`)
:::


## Disclaimer

The information in this report is provided on an "as is" basis and its use does not imply endorsement or verification by United States Environmental Protection Agency (EPA). These data and any interpretations do not necessarily represent the EPA’s official position or viewpoint, expressed or implied. EPA strongly recommends careful attention be paid to metadata files associated with these data to better understand limitations, restrictions or intended use. The U.S. EPA shall not be held liable for improper or incorrect use of the data. Any reference to specific commercial products, processes, or services by service mark, trademark, manufacturer, or otherwise, does not constitute or imply their endorsement, recommendation or favoring by EPA. The EPA seal and logo shall not be used in any manner to imply endorsement of any commercial product or activity by EPA or the United States Government.


## Introduction

This report evaluates quality control data associated with a set of core physical and chemical measures common among CWA106 monitoring programs. Data used in this report were queried from the Water Quality Portal ([WQP](https://www.waterqualitydata.us/){.external target="_blank"}), which collates from a variety of sources including the EPA's Water Quality Exchange (WQX) Data Warehouse. Relevant parameters available in data from the `r allSummary$data$OrganizationFormalName[1]` CWA106 Program include `r included_params`. `r params_not_found_message`

In preparing data for analysis, non-detects (values below detection limit) were treated numerically as half of the reported detection limit (e.g., a value reported as '<1 mg/L' would be treated as '0.5 mg/L'). Estimated values above a calibration curve were treated as equivalent to the maximum known value (e.g., a value reported as '>250 mg/L' would be treated as '250 mg/L'). Data users should address censored values in ways that are appropriate for their motivating questions and in consultation with relevant literature (including EPA's [QA/G-9 Guidance For Data Quality Assessment](https://www.epa.gov/sites/default/files/2015-06/documents/g9-final.pdf){.external target="_blank"}).


[Map of sampling locations]

```{r}
#| fig-cap: Figure 1. Map of sampling locations included in this report.
#| echo: false



```




## Blank Performance


```{r}
#| eval: false
#| include: false
#| fig-cap: Table 1. Blank performance for each analyte in each year, based on data input into the Water Quality Portal.
### https://gt.albert-rapp.de/getting_started.html
### create cleaner sample type 
if (nrow(allSummary$blank_proc) == 0) {
  blank_table <- data.frame('Blanks collected' = c(0))
} else {
  blank_int <- allSummary$blank_summary
  blank_int$`Sample Type` <- sapply(strsplit(blank_int$ActivityTypeCode, '-'), FUN = '[[', 2)
blank_int$`Above MDL (%)` <- paste0(round(blank_int$exceedances / blank_int$n, 3)*100, '%')
blank_int[, 5] <- round(blank_int[, 5], 3)
names(blank_int)[c(3, 5:7, 9)] <- c('Analyte', 'Year', 'Blanks', 'Mean value', 'Above MDL (no.)')
blank_table <- blank_int[, c(10, 3, 5:7, 9, 11)]
blank_table$`Mean value` <- sprintf("%.3f", round(blank_table$`Mean value`,3)) # keep trailing zeros
# allSummary$blank_proc$date <- as.POSIXct(allSummary$blank_proc$ActivityStartDate, format = '%Y-%m-%d')
}
# Summary of blank performance by year:

gt(blank_table)

```



```{r}
#| echo: false
if (nrow(allSummary$blank_proc) == 0) {
  blank_table <- data.frame('Blanks collected' = c(0))
} else {
  blank_int <- plyr::ddply(allSummary$blank_proc, c('ActivityTypeCode', 'CharacteristicName'), plyr::summarise,
                           `Blanks (no.)`            = length(na.omit(year)),
                           `Years with blanks` = length(na.omit(unique(year))),
                           `Blanks above MDL (no.)`  = sum(exceedMDL, na.rm = TRUE),
                           `pct_above_MDL`           = sum(exceedMDL, na.rm = TRUE)/length(na.omit(exceedMDL))*100,
                           `Blanks above MDL (%)`    = paste0(round(pct_above_MDL,0), '%')
  )
  ### get total samples per analyte to calculate blank ratio
  tot_samples <- plyr::ddply(allSummary$data, c('CharacteristicName'), plyr::summarise,
                           `Total observations`            = length(na.omit(ResultMeasureValue))
  )
  blank_int$`Total observations` <- tot_samples$`Total observations`[tot_samples$CharacteristicName %in% blank_int$CharacteristicName]
  blank_freq           <- blank_int$`Blanks (no.)` / blank_int$`Total observations`*100
  blank_int$`Blank frequency`    <- paste0(round(blank_freq, 1), '%')

blank_int$`Sample Type` <- sapply(strsplit(blank_int$ActivityTypeCode, '-'), FUN = '[[', 2)
blank_int$Analyte       <- blank_int$CharacteristicName
blank_table             <- blank_int[, c(10, 11, 3, 5, 4, 7, 9)]
blank_table$`Blank frequency >10%?` <- ifelse(blank_freq >= 10.0, "\u2705", "\u274C")
blank_table$`<5% of blanks contaminated?` <- ifelse(blank_int$pct_above_MDL <= 5.0, "\u2705", "\u274C")
}
```

```{r}
#| echo: false
#| warning: false
#| fig-width: 8
#| fig-cap: Figure 2. Field blank performance for each analyte over time (black). Red points indicate detection limits for each analyte during each sampling event.
if (nrow(allSummary$blank_proc) == 0) {
 ggplot()
} else {
allSummary$blank_proc$date <- as.POSIXct(allSummary$blank_proc$ActivityStartDate, format = '%Y-%m-%d')

ggplot(allSummary$blank_proc[grepl(x = tolower(allSummary$blank_proc$ActivityTypeCode), pattern = 'field blank'), ], aes(x = date, y = ResultMeasureValue)) +
  geom_line(linetype = 2) + geom_point() + 
  facet_wrap(. ~ CharacteristicName, scales = 'free_y') + theme_bw() + theme(plot.margin = margin(r = 40, unit = "pt")) +
  geom_line(aes(y = DetectionQuantitationLimitMeasure.MeasureValue, x = date), col = 'red', linetype = 1) +
  geom_point(aes(y = DetectionQuantitationLimitMeasure.MeasureValue, x = date), col = 'red', size = 0.5) +
  labs(y = 'Blank value', x = '')
}
```


```{r}
#| echo: false
tmp <- table(allSummary$blank_proc$ActivityTypeCode[!is.na(allSummary$blank_proc$ResultMeasureValue)])

### if TN is always a calculated value, exclude it from total samples
if (all((length(unique(allSummary$data[allSummary$data$CharacteristicName == 'TN', 'ResultAnalyticalMethod.MethodIdentifier'])) == 1) & (unique(allSummary$data[allSummary$data$CharacteristicName == 'TN', 'ResultAnalyticalMethod.MethodIdentifier']) == 'Calculated'))) {
  sonde_parameters <- c(sonde_parameters, 'TN')
}

number_of_lab_blanks <- ifelse(any(grepl(x = names(tmp), pattern = 'Lab')), sum(tmp[grepl(x = names(tmp), pattern = 'Lab')]), 0)
number_of_field_blanks <- ifelse(any(grepl(x = names(tmp), pattern = 'Field')), sum(tmp[grepl(x = names(tmp), pattern = 'Field')]), 0)
no_years <- suppressWarnings(length(unique(na.omit(format(allSummary$data$ActivityStartDate, '%Y')))))
### number of samples, excluding sonde data
no_samples <- sum(!is.na(allSummary$data$ResultMeasureValue[!(allSummary$data$CharacteristicName %in% sonde_parameters)]))


### correct for cases where no blank data are reported
if (length(number_of_lab_blanks) == 0)   number_of_lab_blanks   <- 0
if (length(number_of_field_blanks) == 0) number_of_field_blanks <- 0 
if (nrow(allSummary$blank_summary) == 0)  {
  number_field_blank_detects  <- NA
  percent_field_blank_detects <- NA
} else {
  number_field_blank_detects <- sum(allSummary$blank_summary$exceedances[grep(x =   allSummary$blank_summary$ActivityTypeCode, pattern = 'Field Blank')], na.rm = TRUE)
  
  percent_field_blank_detects <- round(sum(allSummary$blank_summary$exceedances[grep(x = allSummary$blank_summary$ActivityTypeCode, pattern = 'Field Blank')], na.rm = TRUE) / sum(allSummary$blank_summary$n[grep(x = allSummary$blank_summary$ActivityTypeCode, pattern = 'Field Blank')], na.rm = TRUE), 3)*100
  }



### identify parameters without QC data (exclude sonde measurements)
parameters_without_blanks <- unique(allSummary$data$CharacteristicName)[!(unique(allSummary$data$CharacteristicName) %in% c(unique(allSummary$blank_summary$CharacteristicName), sonde_parameters))]

if(length(parameters_without_blanks) < 1) {
 parameter_blank_message <- 'All of the laboratory parameters included in this report had some blank data.'
} else {
  parameter_blank_message <- paste0('Laboratory analytes without blanks included ', paste0(parameters_without_blanks, collapse = ', '),'.')
}

### identify years without QC data 
### can make this more complex, identifying params without blank data in years when other params had reported blank data
years_with_blanks    <- as.numeric(unique(allSummary$blank_summary$year))
years_with_data      <- sort(as.numeric(na.omit(unique(substr(allSummary$data$ActivityStartDateTime, 1, 4)))))
years_without_blanks <- years_with_data[!(years_with_data %in% years_with_blanks)]

if(length(years_without_blanks) < 1) {
 year_blank_message <- 'Blank data was reported in every year during the period of record covered by this report.' 
} else {
  year_blank_message <- paste0('Blank data was not reported during the following years: ', paste0(years_without_blanks, collapse = ', '),'.')
}

```

There were `r sum(allSummary$blank_summary$Blanks, na.rm = TRUE)` lab and field blanks collected during the period of interest. This total number is composed of `r number_of_lab_blanks` laboratory blanks and `r number_of_field_blanks` field blanks. Of the field blanks, `r number_field_blank_detects` (`r percent_field_blank_detects`%) had detectable analyte levels.

`r year_blank_message`

`r parameter_blank_message`


```{r}
#| echo: false
#| fig-cap: Table 1. Field blank performance for each analyte. Rightmost two columns assess whether the project's blank collection rates and blank performance meet target criteria (green checks) or not (red 'X' marks).
gt(blank_table)
```



### Recommendations: Blanks

```{r}
#| include: false
### recommendations based on blank data 
rec_list <- list()
if (number_of_lab_blanks == 0) {
  lab_blank_rec <- '**Laboratory blanks** were not present in data found in the Water Quality Portal. These data appear to have been either not collected or not input into WQP. More comprehensive collection and reporting of laboratory blanks for conducting meaningful assessments of data quality, identifying contamination, and making comparisons among data from different laboratories.'
} else if (number_of_lab_blanks < no_years) {
  lab_blank_rec <- '**Laboratory blanks** appear in the reported data at a rate of less than one lab blank per year, indicating either minimal collection or incomplete reporting of these QC data to WQP. More comprehensive reporting of laboratory blanks into WQP would enable more meaningful assessment of data quality differences among datasets and laboratories.'
} else if (number_of_lab_blanks < 0.10*no_samples) {
  lab_blank_rec <- paste0('**Laboratory blanks** appear in the data reported to WQP, but represent just ', round(number_of_lab_blanks / no_samples*100), '% of all samples. Recommendations include ensuring that laboratory blanks are reported comprehensively and increasing the frequency of laboratory blanks to at least 10% of lab samples.')
} else if (number_of_lab_blanks >= 0.10*no_samples) {
  lab_blank_rec <- '**Laboratory blanks** represent at least 10% of all lab samples. Project Managers are commended for their high level of effort and reporting on this quality metric.'
}
rec_list$lab_blank <- lab_blank_rec


### field blanks
if (number_of_field_blanks == 0) {
  field_blank_rec <- '**Field blanks** were either not collected or not input into WQP. More comprehensive reporting of field blanks in WQX is critical for evaluating contamination and bias in the generated data.'
} else if (number_of_field_blanks < 0.10*no_samples) {
  field_blank_rec <- paste0('**Field blanks** appear in the reported data at a rate of ', round(number_of_field_blanks / no_samples*100), '% of all relevant samples (excluding sonde data). This indicates either limited collection or incomplete reporting of these important QC data to WQP. Increasing the frequency of field blanks to at least 10% is highly recommended. Improving collection and reporting of field blank data is critical to ensure that project data can be useful for decision-making.')
} else if (number_of_field_blanks >= 0.10*no_samples) {
  field_blank_rec <- paste0('**Field blanks** represent ', round(number_of_field_blanks / no_samples*100), '% of all samples (excluding sonde data), above the target level of 10%. Project Managers are commended for their high level of effort and reporting on this quality metric.')
}
rec_list$field_blank <- field_blank_rec

### Inconsistent blank collection (over time and among parameters)
if ((length(years_without_blanks) == 0)) {
  blank_collection_consistency_time <- '**Blank collection over time:** Blanks were collected and reported in every year, which merits recognition. '
} else if ((length(years_without_blanks) > 0)) {
  blank_collection_consistency_time <- '**Blank collection over time:** Blanks were reported inconsistently over time. Some years had no blank data reported. '
}
if ((length(parameters_without_blanks) == 0)) {
  blank_collection_consistency_param <- '**Blank collection by parameter:** Evaluating consistency in blank reporting across parameters, it is observed that the reported data include at least one blank for every laboratory analyte. Project managers should ensure that this comprehensive level of quality control data continues.'
} else if ((length(parameters_without_blanks) > 0)) {
  blank_collection_consistency_param <- '**Blank collection by parameter:** Blanks were not reported consistently across parameters, with some analytes having no blank data found in the Water Quality Portal. Project managers should ensure that every analyte has field and lab blank data reported for every year in which that analyte is measured.'
}

### blank performance and possible sample handling issues
if(is.na(percent_field_blank_detects)) {
  blank_issues <- 'The absence of blank data for core CWA106 parameters makes it impossible to evaluate bias in the dataset.'
} else if(percent_field_blank_detects <= 10) {
  blank_issues <- 'This is encouraging and indicates consistent attention to detail during sample collection and handling.'
} else {
  blank_issues <- 'This magnitude of contamination deserves further attention and effort to identify its causes. Increasing the frequency of blank collection and re-training staff are among the possible ways to reduce contamination during sample collection and handling.'
}

# ### simplify message if there's no blank data
# if (number_of_field_blanks == 0) {

rec_list$blank_freq_time  <- blank_collection_consistency_time
rec_list$blank_freq_param <- blank_collection_consistency_param
rec_list$blank_issues     <- paste0('**Blank performance:** Overall, ', percent_field_blank_detects, '% of blanks had evidence of contamination. ' , blank_issues, ' See Table 1 for parameter-specific performance and to identify specific areas for improvement.')
```

-   `r rec_list$lab_blank`

-   `r rec_list$field_blank`

-   `r rec_list$blank_freq_time`

-   `r rec_list$blank_freq_param`

-   `r rec_list$blank_issues`



## Replicate Performance

<!-- `r sum(allSummary$blank_summary$Blanks, na.rm = TRUE)` lab and field blanks appeared in the dataset: `r number_of_lab_blanks` laboratory blanks and `r number_of_field_blanks` field blanks. Of the field blanks, `r number_field_blank_detects` (`r percent_field_blank_detects`%) had detectable analyte levels. -->

```{r}
#| echo: false
#| warning: false
#| fig-width: 8
#| fig-cap: Figure 3. Replicate performance by parameter during the period of record for this report. Each point shows variation among a set of simultaneously-collected replicates, overlaid on boxplots indicating the 25th, 50th (bar), and 75th percentiles of the distribution for each parameter. Boxplot whiskers extend to the largest value up to 1.5 times the interquartile range beyond the 25th and 75th percentiles. The total number of replicate sets are indicated above each parameter. A dashed horizontal line is shown at the 20% level for reference.
### consider adding 95% CIs (notch = TRUE looks bad): https://stackoverflow.com/a/22200586
### TODO: annotate with total number from 
rep_totals_by_param <- data.frame(table(allSummary$rep_proc$CharacteristicName))
rep_totals_by_param$tot_samples <- tot_samples$`Total observations`[tot_samples$CharacteristicName %in% as.character(rep_totals_by_param$Var1)]
rep_totals_by_param$label_text <- paste0(as.character(rep_totals_by_param$Freq))#, 
                                         # '\n', round(rep_totals_by_param$Freq/ rep_totals_by_param$tot_samples * 100), '%')
rep_totals_by_param$label_text[1] <- paste0('n = ', rep_totals_by_param$label_text[1])

label_location <- as.numeric(quantile(x = allSummary$rep_proc$RPD, probs = 0.95, na.rm = TRUE))

ggplot(allSummary$rep_proc[grepl(x = allSummary$rep_proc$ActivityTypeCode, pattern = 'Field replicate'), ], aes(x = CharacteristicName, y = RPD)) +
  geom_boxplot(outlier.shape = NA) + theme_bw() + #  theme(axis.text.x=element_blank())
  ylab(allSummary$rep_summary$variation_measure[1]) + xlab('') + scale_y_continuous(labels = function(x) paste0(x, "%")) +  
  geom_jitter(width=0.2, size=2, alpha = 0.2) +
  geom_text(
    data = rep_totals_by_param,
    aes(x = Var1, y = label_location,label = label_text),
    vjust = -0.5,
    size = 4,
    stat = "identity"
  ) +
  geom_hline(yintercept = 20, linetype = 2)


### number of field and lab reps
total_reps <- sum(allSummary$rep_summary$n, na.rm = TRUE)
number_of_field_reps <- sum(allSummary$rep_summary$n[grep(x = allSummary$rep_summary$ActivityTypeCode, pattern = 'Field')], na.rm = TRUE)
number_of_lab_reps   <- sum(allSummary$rep_summary$n[grep(x = allSummary$rep_summary$ActivityTypeCode, pattern = 'Lab')], na.rm = TRUE)
samples_for_rep_comparison <- sum(!is.na(allSummary$data$ResultMeasureValue[allSummary$data$CharacteristicName %in% allSummary$rep_summary$CharacteristicName])) # only use params that have field or lab reps

### Replicate sets above 20% variation, by parameter
### can split this into field vs. lab in future if orgs report lab reps
above20_by_param <- plyr::ddply(allSummary$rep_proc, c('CharacteristicName'), plyr::summarize,
      `Number of replicate sets`       = sum(!is.na(RPD)),
      `Sets with variation above 20%` = sum(RPD > 20, na.rm = TRUE),
      `Proportion above 20%`     = paste0(round(100*(sum(RPD > 20, na.rm = TRUE) / sum(!is.na(RPD)))), '%')
      )
above20_by_param$tot_samples <- tot_samples$`Total observations`[tot_samples$CharacteristicName %in% above20_by_param$CharacteristicName]
above20_by_param$rep_freq                  <- above20_by_param$`Number of replicate sets`/ above20_by_param$tot_samples * 100
above20_by_param$`Replicate frequency (%)` <- paste0(round(above20_by_param$rep_freq, 0), '%')
above20_by_param$`Replicate frequency >10%?`       <- ifelse(above20_by_param$rep_freq >= 10.0, "\u2705", "\u274C")
above20_by_param$`Are <10% of reps high-variation?` <- ifelse(above20_by_param$`Proportion above 20%` <= 10.0, "\u2705", "\u274C")

names(above20_by_param)[grep(x = names(above20_by_param), pattern = 'CharacteristicName')] <- 'Analyte'

### parameters without reps (even sonde data should be replicated)
parameters_without_reps <- unique(allSummary$data$CharacteristicName)[!(unique(allSummary$data$CharacteristicName) %in% c(unique(allSummary$rep_summary$CharacteristicName)))]

if(length(parameters_without_reps) < 1) {
 parameter_rep_message <- 'All of the laboratory parameters included in this report had some replicate data.'
} else {
  parameter_rep_message <- paste0('Laboratory analytes without replicates included ', paste0(parameters_without_reps, collapse = ', '),'.')
}
TN_note <- NULL
### add a note if TN is a calculated quantity
if (('TN' %in% parameters_without_reps) && (unique(allSummary$data[allSummary$data$CharacteristicName == 'TN', 'ResultAnalyticalMethod.MethodIdentifier']) == 'Calculated')) {
  TN_note <- 'Although TN is identified as a calculated parameter in this study, replicate values should be calculated and reported in WQP. Replicates can be calculated using replicate data for the component compounds (TKN-N, NH3-N, NO3+NO2-N).'
}

### identify years without reps
### can make this more complex, identifying params without blank data in years when other params had reported blank data
years_with_reps    <- as.numeric(unique(allSummary$rep_summary$year))
years_with_data    <- as.numeric(na.omit(unique(substr(allSummary$data$ActivityStartDateTime, 1, 4))))
years_without_reps <- years_with_data[!(years_with_data %in% years_with_reps)]

if(length(years_without_reps) < 1) {
 year_rep_message <- 'Replicate data was reported in every year during the period of record covered by this report.' 
} else {
  year_rep_message <- paste0('Replicate data was not reported during the following years: ', paste0(years_without_reps, collapse = ', '),'.')
}

### identify the three params with the highest variation
median_vals <- plyr::ddply(allSummary$rep_proc, c('CharacteristicName'), plyr::summarize, 
                           medianRPD = median(RPD, na.rm = TRUE)
                           )
highest_variation_params <- median_vals[order(median_vals$medianRPD, decreasing = TRUE), ][1:3, ]
median_variation_above20 <- median_vals$CharacteristicName[median_vals$medianRPD > 20]

total_reps_above20      <- sum(allSummary$rep_proc$RPD > 20, na.rm = TRUE)
reps_pct_above20        <- round(total_reps_above20 / sum(!is.na(allSummary$rep_proc$RPD)) * 100)
summary_reps_above20    <- paste0(total_reps_above20, ' sets of replicates (', reps_pct_above20, '% of the total) had ', tolower(allSummary$rep_summary$variation_measure[1]), ' exceeding the target 20%.')

variation_measure_message <- ifelse(allSummary$rep_summary$variation_measure[1] == "Coefficient of Variation", 
                                    paste0('Some replicates sets included as many as ', max(table(allSummary$rep_raw$id)), ' individual replicates. This could indicate errors during data entry into WQP, or it may reflect exceptional levels of replication. This report could not resolve those uncertainties, and so used the coefficent of variation (CV) as the measure of variation among replicates. Lower CVs indicate less variation among replicates.'), 
                                    'Replicate sets contained a maximum of two individual samples, and so the relative percent difference (RPD) was used as the measure of variation among replicates. Lower RPDs indicate lower levels of variation among replicates.')

```

Over the period of record used for this report, `r allSummary$data$OrganizationFormalName[1]` collected `r total_reps` sets of replicates. This combined total includes `r number_of_field_reps` field replicates and `r number_of_lab_reps` lab replicates. Field replicates comprised `r round(number_of_field_reps/ samples_for_rep_comparison*100)`% of all samples.

In comparison to the 20% target level of variation, replicate performance was `r ifelse(reps_pct_above20 <= 10, 'excellent', ifelse(reps_pct_above20 <= 25, 'decent', ifelse(reps_pct_above20 <= 50, 'not good', ifelse(reps_pct_above20 > 50, 'poor'))))`. `r summary_reps_above20`

`r variation_measure_message`

`r year_rep_message`

`r parameter_rep_message`

```{r}
#| echo: false
#| warning: false
#| label: Table2
#| fig-cap: Table 2. Summary of replicate collection rates and performance by analyte. The rightmost two columns assess whether the project's collection rates and replicate performance meet target criteria (green checks) or not (red 'X' marks).
gt(above20_by_param[, c(1:3, 7, 4, 8:9)])
```

Replicate variation is expected to be elevated for analytes that are not homogenous throughout the water column. This group includes *E. coli*, turbidity/total suspended solids (TSS), total Kjeldahl nitrogen (TKN), total phosphorus, and other analytes that rely on particulate fractions. This variation itself may be inconsistent over time (Figure 4), varying as a function of discharge and local in-stream deposition/erosion dynamics.

Analytes that measure dissolved quantities (temperature, DO, ions) are expected to be more consistent between replicate samples, and variation in these analytes is more likely to reflect discrepancies in sample collection, handling, and contamination. However, if dissolved analytes (NO3-N, NH3-N) are filtered in the lab rather than in the field, there are opportunities for compounds to sorb or desorb from particles in the collected water, providing another source of variation captured by replicates.

```{r}
#| echo: false
#| warning: false
#| fig-width: 8
#| fig-cap: Figure 4. Field replicate performance over time for each parameter. Note that y-axis scales vary across facets. A dashed horizontal line is shown in each facet at the 20% level for reference.

allSummary$rep_proc$date <- as.POSIXct(sapply(X = strsplit(allSummary$rep_proc$id, '__'), FUN = '[[', 6), format = '%Y-%m-%d')

ggplot(allSummary$rep_proc[grepl(x = allSummary$rep_proc$ActivityTypeCode, pattern = 'Field replicate'), ], aes(x = date, y = RPD)) +
  geom_line(linetype = 2) + geom_point() + scale_y_continuous(labels = function(x) paste0(x, "%")) +
  facet_wrap(. ~ CharacteristicName, scales = 'free_y') + theme_bw() +
  geom_hline(yintercept = 20, linetype = 3) +
  labs(y = allSummary$rep_summary$variation_measure[1], x = '') # '(\u00b11SD)'

```


### Recommendations: Replicates

```{r}
#| include: false
### recommendations based on replicate data 
if (number_of_lab_reps == 0) {
  lab_rep_rec <- '**Laboratory replicates** were not present in data found in the Water Quality Portal. These data appear to have been either not collected or not input into WQP. More comprehensive reporting of laboratory replicates into WQP would enable more meaningful assessment of data quality.'
} else if (number_of_lab_reps < 0.10*samples_for_rep_comparison) {
  lab_rep_rec <- paste0('**Laboratory replicates** appear in the data reported to WQP, but represent just ', round(number_of_lab_reps / samples_for_rep_comparison*100), '% of all samples. Recommendations include ensuring that laboratory replicates are reported comprehensively and increasing the frequency of laboratory replicates to at least 10% of lab samples.')
} else if (number_of_lab_reps >= 0.10*samples_for_rep_comparison) {
  lab_rep_rec <- '**Laboratory replicates** represent at least 10% of all lab samples. Project Managers are commended for their high level of effort and reporting on this quality metric.'
}
rec_list$lab_rep <- lab_rep_rec


### field reps
if (number_of_field_reps == 0) {
  field_rep_rec <- '**Field replicates** were either not collected or not input into WQP. More comprehensive reporting of field replicates in WQX is critical for evaluating contamination and bias in the generated data.'
} else if (number_of_field_reps < 0.10*samples_for_rep_comparison) {
  field_rep_rec <- paste0('**Field replicates** appear in the data reported to WQP at a rate of ', round(number_of_field_reps / samples_for_rep_comparison*100), '% of all samples. Recommendations include ensuring that field replicates are reported comprehensively and increasing the frequency of field replicates to at least 10% of all samples.')
}else if (number_of_field_reps >= 0.10*samples_for_rep_comparison) {
  field_rep_rec <- paste0('**Field replicates** represent ', round(number_of_field_reps / samples_for_rep_comparison*100), '% of all samples (excluding sonde data), above the target level of 10%. Project Managers are commended for their high level of effort and reporting on this quality metric.')
}
rec_list$field_rep <- field_rep_rec

### Inconsistent rep collection (over time and among parameters)
if ((length(years_without_reps) == 0)) {
  rep_collection_consistency_time <- '**Replicate collection over time:** Replicates were collected and reported in every year, which merits recognition. '
} else if ((length(years_without_reps) > 0)) {
  rep_collection_consistency_time <- '**Replicate collection over time:** Replicates were reported inconsistently over time, with some years having no replicates reported. '
}
if ((length(parameters_without_reps) == 0)) {
  rep_collection_consistency_param <- '**Replicate collection by parameter:** Evaluating consistency in replicate reporting across parameters, it is observed that the reported data include at least one replicate for every laboratory analyte. Project managers should ensure that this comprehensive level of quality control data continues.'
} else if ((length(parameters_without_reps) > 0)) {
  rep_collection_consistency_param <- paste0('**Replicate collection by parameter:** Replicates were not reported consistently across parameters, with some analytes having no replicate data found in the Water Quality Portal. Project managers should ensure that every analyte has field and lab replicate data reported for every year in which that analyte is measured. ', parameter_rep_message)
}

### replicate performance and possible sample handling issues
if(reps_pct_above20 <= 10) {
  rep_issues <- 'This is encouraging and suggests consistency and attention to detail during sample collection and handling.'
} else {
  rep_issues <- 'This level of imprecision suggests examining opportunities for improved consistency and attention to detail during sample collection and handling. Increasing the frequency of replicates is recommended to help identify the underlying issues.'
}


rec_list$rep_freq_time  <- rep_collection_consistency_time
rec_list$rep_freq_param <- rep_collection_consistency_param
rec_list$rep_issues     <- paste0('**Replicate performance:** Overall, ', reps_pct_above20, '% of replicates had problematic levels of imprecision. ' , rep_issues, ' See Table 2 and Figure 4 for parameter-specific performance and to identify specific areas for improvement.')

```

-   `r rec_list$lab_rep`

-   `r rec_list$field_rep`

-   `r rec_list$rep_freq_time`

-   `r rec_list$rep_freq_param` `r if(!is.null(TN_note)) TN_note`

-   `r rec_list$rep_issues`



SUPPORTING_DATA_TEXT

```{r}
#| echo: false
#| warning: false
blank_file <- tempfile(fileext = '.csv')
blank_data <- write.csv(allSummary$blank_proc, file = blank_file, row.names = FALSE)
appendToFileName <- paste0("_", allSummary$data$OrganizationIdentifier[1], "_", min(years_with_data, na.rm = TRUE), "_", max(years_with_data, na.rm = TRUE))
### can we avoid using downloadthis package?
downloadthis::download_file(path = blank_file,
  output_name  = paste0("Blank_data_raw", appendToFileName),
  button_label = "Download blank data", 
  button_type  = "default",
  has_icon     = TRUE,
  icon         = "fa fa-save"
  )
```

\n 

```{r}
#| echo: false
#| warning: false
blank_summary_file <- tempfile(fileext = '.csv')
blank_summary_data <- write.csv(allSummary$blank_summary, file = blank_summary_file, row.names = FALSE)
downloadthis::download_file(path = blank_summary_file,
  output_name  = paste0("Blank_data_summary", appendToFileName),
  button_label = "Download blank data (annual summary)",
  button_type  = "default",
  has_icon     = TRUE,
  icon         = "fa fa-save"
  )
```

```{r}
#| echo: false
#| warning: false
rep_file <- tempfile(fileext = '.csv')
rep_data <- write.csv(allSummary$rep_raw, file = rep_file, row.names = FALSE)
downloadthis::download_file(path = rep_file,
  output_name  = paste0("Replicate_data_raw", appendToFileName),
  button_label = "Download replicate data (raw)",
  button_type  = "default",
  has_icon     = TRUE,
  icon         = "fa fa-save"
  )
```
    
    
```{r}
#| echo: false
#| warning: false
rep_file2 <- tempfile(fileext = '.csv')
rep_data2 <- write.csv(allSummary$rep_proc, file = rep_file2, row.names = FALSE)
downloadthis::download_file(path = rep_file2,
  output_name  = paste0("Replicate_data_processed", appendToFileName),
  button_label = "Download replicate data (processed)",
  button_type  = "default",
  has_icon     = TRUE,
  icon         = "fa fa-save"
  )
```
    
```{r}
#| echo: false
#| warning: false
rep_summary_file <- tempfile(fileext = '.csv')
rep_summary_data <- write.csv(allSummary$rep_summary, file = rep_summary_file, row.names = FALSE)
downloadthis::download_file(path = rep_summary_file,
  output_name  = paste0("Replicate_data_summary", appendToFileName),
  button_label = "Download replicate data (annual summary)",
  button_type  = "default",
  has_icon     = TRUE,
  icon         = "fa fa-save"
  )
```


```{r}
#| echo: false
#| warning: false
data_file <- tempfile(fileext = '.csv')
data_summary_data <- write.csv(allSummary$data, file = data_file, row.names = FALSE)
downloadthis::download_file(path = data_file,
  output_name  = paste0("Processed_data", appendToFileName),
  button_label = "Download post-QC data",
  button_type  = "default",
  has_icon     = TRUE,
  icon         = "fa fa-save"
  )
```


```{r}
#| echo: false
#| warning: false
raw_data_file <- tempfile(fileext = '.csv')
raw_data_summary_data <- write.csv(tribeData2[order(tribeData2$OrganizationFormalName, tribeData2$ActivityStartDate, tribeData2$MonitoringLocationIdentifier, tribeData2$CharacteristicName), ], file = raw_data_file, row.names = FALSE)
downloadthis::download_file(path = raw_data_file,
  output_name  = paste0("WQP_data", appendToFileName),
  button_label = "Download WQP data",
  button_type  = "default",
  has_icon     = TRUE,
  icon         = "fa fa-save"
  )
```
